Does this Star Belong? Data-Driven Characterization of Stellar Streams \\ with Mixture Density Networks

### 2023-11-22 src/tex/ms.tex
fix calcs
Signed-off-by: nstarman <nstarman@users.noreply.github.com>
[showyourwork] overleaf sync
Signed-off-by: nstarman <nstarman@users.noreply.github.com>

\documentclass[twocolumn]{aastex631}
\usepackage{showyourwork}
\usepackage{amsmath, amssymb}
\usepackage{gensymb}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{dsfont}
\usepackage[inline]{enumitem}
\usepackage{cleveref}
\usepackage{todonotes}

% Change the section names
\renewcommand{\sectionautorefname}{\S\!}%
\renewcommand{\subsectionautorefname}{\S\!\!}%
\renewcommand{\subsubsectionautorefname}{\S\!}%
\renewcommand{\equationautorefname}{Eq.}
\renewcommand{\tableautorefname}{Table}

\newcommand{\code}[1]{\textsc{#1}}
\newcommand{\package}[1]{\code{#1}}
\newcommand{\stream}[1]{#1}
\newcommand{\dataarchive}[1]{\texttt{#1}}
\newcommand{\Gaia}{\dataarchive{Gaia}}
\newcommand{\PanStarrs}{\dataarchive{Pan-STARRS1}}

@@

% Statistics
\newcommand{\pdf}{\mcal{P}}
\newcommand{\cdf}{\mfk{P}}
\newcommand{\likelihood}{\mcal{L}}
\newcommand{\prior}{\mcal{\pi}}
\newcommand{\posterior}{p}

@@

@@

% Title
\title{Characterizing Stellar Streams by ML}

% Author list
\author{Nathaniel Starkman \& Jacob Nibauer}
\author{Jo Bovy}
\author{Jeremy Webb}
\author{Ana Bonaca}

% Abstract with filler text
\begin{abstract}
    Stellar stream membership likelihoods with ML. It works!
\end{abstract}

@@

    Stellar streams are the disrupted remnants of globular clusters and
    satellite galaxies, and their census has grown extensively in the Milky Way
    thanks to all-sky astrometric missions like {\it Gaia} and low-surface
    brightness photometric surveys (e.g., \citealt{2018ApJ...862..114S,
    2021ApJ...914..123I}).  The formation of streams can be traced back to a
    progenitor cluster, whose stars are gradually lost to the tidal field of the
    host galaxy and extend along a series of similar orbits producing tidal
    tails, which can sometimes span several tens of degrees across the sky. The
    orbital proximity of stars belonging to a stream makes them sensitive
    tracers of the mass distribution of galaxies, sourced by both the baryonic
    and dark matter components.

    The average properties of a stream (e.g., the mean position and velocity
    track of the tidal tails) can be used to reconstruct the global dark matter
    distribution of the stellar halo (e.g., \JN{cite}), while local variations
    in the stream track and surface density encode information about the
    stream's encounters with baryonic structures such as (e.g.) the stellar bar
    (\JN{cite}), and non-luminous structures like dark matter subhalos
    (\JN{cite}). Importantly, these galactic components span orders of magnitude
    in mass and spatial extent, highlighting the sensitivity of streams in
    constraining the structure of the galaxy.

    While several methods have been developed to model stream properties as a
    function of global halo characteristics (e.g., flattening and
    orientation;\JN{cite}) and local substructures (e.g., the bar's pattern
    speed and the mass of subhalo perturbers; \JN{cite}), these methods often
    rely on the existence of a homogenized stream-tracer population that
    characterizes a given stream across each of the observable phase-space
    dimensions. Additionally, because even the fine-grain properties of streams
    are influenced by the detailed mass distribution of its host galaxy, it is
    important to propagate errors both in measurement uncertainty and membership
    uncertainty (i.e., which stars belong to the stream versus the background)
    when attempting to model a stream. Otherwise, a poorly characterized stream
    could bias constraints on (e.g.) the total mass of the galaxy, the dark
    matter halo shape, and the mass of subhalo perturbers. In order to generate
    more precise constraints on the structure of the galaxy, it is therefore
    imperative that a statistically sound and homogenized catalog of stellar
    streams is produced.

    While streams are well described in (e.g.) action angle coordinates or
    $N-$body simulations, these methods rely on prior assumptions about the
    galaxy and its underlying gravitational potential. Therefore, it is
    necessary to devise data-driven methods to characterize stellar streams
    without appealing to strong prior assumptions about the galaxy's structure.

    There are a few previous works which have modeled stellar streams in this
    context. \citet{Patrick+2022} developed a data-driven spline-based method
    for modeling the photometric properties and on-sky positions of previously
    identified streams. Their method works by fitting the stream's stellar
    surface density with a series of splines, capturing both surface brightness
    variations along the stream and changes in the position of the stream track
    and width. Their method also utilize a simultaneous isochrone fit to a given
    stream, from which a distance modulus can be derived. While this method is
    very successful at characterizing the photometric properties of a stream, it
    does not consider the kinematic dimensions which have become increasingly
    well-measured with successive data releases from {\it Gaia}, as well as
    targeted observations. From a population of RR Lyrae stars,
    \citet{Price-Whelan+2019} utilized a mixture model to fit the globular
    cluster stream Pal-5 in the space of proper motions and heliocentric
    distances. The mixture modeling approach enables a simultaneous fit to both
    the stream and background density, allowing for better separation of the
    stream from the background distribution. However, the track of the stream in
    on-sky coordinates is not fit. The method identified 27 RR Lyrae stars
    consistent with being members of the stream. While useful for determining
    the average properties of a stream (e.g., its mean distance track), a
    limited population of RR Lyrae stars are not sufficient to capture the
    small-scale properties of streams expected to inform constraints on (e.g.) a
    population of perturbers in the galaxy. \JN{are there other methods?}

    Other methods for extracting stellar streams from the data include
    matched-filter approaches, where over-densities in color-magnitude space are
    inspected as possible clusters undergoing tidal stripping \JN{cite the many
    key papers using this method}. While the matched-filter technique has been
    extremely successful in stream discovery, it is not equipped to characterize
    streams at the level of detailed necessary to inform precise constraints on
    the potential of the galaxy and its substructure.

    In order to address the need for a flexible model of stellar streams both in
    photometric and kinematic spaces, in this work we develop a new method for
    characterizing streams and quantifying the membership probability of
    possible stream stars. Our approach provides several key advantage over
    previous stream modeling efforts, specifically in its ability to jointly
    model the kinematic and photometric properties of streams simultaneously.
    Furthermore, our method is well-suited to fitting streams with only partial
    or incomplete phase-space or photometric observations, enabling a full
    characterization of streams insofar as the available data allows.

    % \begin{quotation}

    %     A stellar stream is the association of stars tidally stripped from a
    %     common progenitor system that orbits in an underlying galactic
    %     gravitational field \citep{Johnston1998, HelmiWhite1999}. Stream
    %     progenitors are gravitationally bound systems, such as satellite dwarf
    %     galaxies or star clusters \citep{Odenkirchen+2001, Majewski+2003}. In
    %     the absence of external interactions these systems are stable, though
    %     still capable of mass loss driven by internal processes
    %     \citep{Hills1975, HeggieHut2003}, e.g. relaxation and binary
    %     interactions. Considering external interactions, these progenitor
    %     systems may be disrupted: completely, as in the merger events that
    %     formed the Galactic stellar halo \citep{Lynden-Bell1967,
    %     SearleZinn1978}; or partially, such as tidal disruption forming a
    %     stellar stream.

    %     In the approximation of a smooth galactic gravitational field, the
    %     progenitor's extent is limited to the tidal radius. Through processes
    %     internal and external, progenitor stars can achieve sufficient energy to
    %     bring them beyond the tidal boundary, thus ``escaping'' into the
    %     galactic field. Escape primarily occurs at both an inner and outer
    %     saddle point of the combined potential -- e.g. the L2 and L3 Lagrange
    %     points for a spherical galactic field \citep{Ross+1997,
    %     FukushigeHeggie2000, BinneyTremaine2008}. Generally, the initial phase
    %     space position of the escaped star is very similar to that of the
    %     progenitor: the configuration position is the tidal boundary and the
    %     star has a low peculiar velocity. Since both the escaped star and its
    %     progenitor orbit in the same underlying galactic potential and have
    %     similar phase space positions their orbits are similar
    %     \citep{BinneyTremaine2008}. Given sufficient time the phase space
    %     separation will increase and an escaped star will lead or trail the
    %     position of the progenitor, depending on the location from which it
    %     escaped. Therefore, a collection of tidally stripped stars will form
    %     extended structures, aka tails or streams, that are few to thousands of
    %     parsecs in length \citep{Johnston1998, HelmiWhite1999, Bovy2014}.

    %     To be a coherent, observable structure the peculiar velocity dispersion
    %     of escaped stars relative to the progenitor must be relatively low, but
    %     does vary from system to system. Stars in kinematically hot streams
    %     (e.g. from progenitors with larger kinematic dispersions) have more
    %     dissimilar orbits from each other and the progenitor than do stars in
    %     cold streams, with smaller dispersions \citep{Johnston1998,
    %     HendelJohnston2015}.  The orbit of a star depends on its phase space
    %     position and the gravitational potential; therefore measuring a star's
    %     position and velocity and observing its orbit reveals the local
    %     potential, which is normally dominated by the Galaxy (primarily the dark
    %     matter halo at large distances) \citep{Gibbons+2014}. However, an orbit
    %     is impractical to observe on human timescales. Cold streams offer a
    %     workaround.  Like measuring one star at many points in one orbit over an
    %     extended period, streams are many stars measured at different points
    %     along approximately one orbit at the same time. Therefore stellar
    %     streams are one of the most promising means to study the Galactic
    %     gravitational potential: the stellar components as well as the dark
    %     matter \citep{Johnston2016}.

    %     On the observational front there has been a large concerted effort in
    %     the field to find new stellar streams \citep[for an extensive list of
    %     known Galactic streams see][]{Mateu2022}, extend the detected extents of
    %     known streams \citep[e.g. for \stream{Palomar 5} alone:][]{Rockosi+2002,
    %     Ibata+2017, Grillmair+2006, Carlberg+2012, Starkman+2019}, and increase
    %     the general purity of stream catalogs.

    % \end{quotation}

    % \todo[inline]{--- Rephrase}

    % \paragraph{one orbit and a single population}

    %     Stellar streams consist of stars that are tidally disrupted from a
    %     progenitor system.  To a first approximation, the stars occupy a single
    %     orbit, that of their progenitor.  Additionally, as stars formed in the
    %     progenitor they retain its photometric and chemical characteristics:
    %     which for globular clusters means following an isochrone.  In this
    %     simplistic picture, astrometrically characterizing a stream requires
    %     only finding it's orbit. Member stars fall directly on the orbit and the
    %     isochrone, making membership likelihood determination trivial.

    %     % Observational errors complicate this picture. Instead of a certain
    %     % membership calculation, we must build a probabilistic model, assigning
    %     % a likelihood that an observed star falls along the stream's orbit.

    % \paragraph{similar orbits and a somewhat noisy population}

    %     Occupying a single orbit implies that streams remain a point in action
    %     space, and having a common origin would thus not become extended
    %     structure. This is not true.  Differences in stream stars' orbits from
    %     the progenitor's arise from various factors including internal dynamics
    %     of the progenitor and the host Galactic potential.

    %     An intrinsic width, in configuration and kinematic space.

    %     Centered on a track, not an orbit.

    %     This defines an astrometric phase-space hyper-tube \cite{STREAMFINDER},
    %     with a distribution radial to the track.

    %     Instead of characterizing the stream by its orbit, we characterize it by
    %     it's track and the distribution around that track.  Computing membership
    %     likelihoods now requires fitting the track and distribution. This cannot
    %     be done without then also fitting the field of stars in which the stream
    %     lives, since the stream and field cannot be clearly separated.

    % \paragraph{gaps orbits and a somewhat noisy population}

    %     This is still a simplistic picture of streams.

    %     Perturbations to the Galactic potential, in the form of time evolution
    %     and non-axisymmetry, as well as substructure can disrupt the stellar
    %     stream.  Even most perturbations don't ruin the idea of the phase-space
    %     hypertube. They either drop the mixture weight (rephrase, not yet
    %     introduced) to 0 or change the width of the hypertube.

    % \paragraph{Review of detection methods}

    %     Identifying streams of globular of clusters is a challenging task.
    %     Streams are very low signal-to-noise structures spread over
    %     comparatively large regions of the sky. They live in a vast background
    %     of field stars.

    %     Historically streams were detected by matched-filter techniques
    %     % Add citations
    %     You took an isochrone and shifted it by a distance modulus until a
    %     stream stood out in longitude and latitude as an extended overdensity
    %     structure.

    %     With the advent of Gaia, we gained access to detailed astrometric data.
    %     For most streams this

    %     poses a challenging task due to their intricate and diffuse nature from
    %     large-scale surveys is challenging due to the complex and overlapping
    %     distributions of stars in different regions of the sky; disentangling
    %     members from the vast background of field stars

    %     \begin{enumerate}
    %         \item Streamfinder \citep{STREAMFINDER}. This method requires the
    %         potential
    %         \item VIA MACHINAE \citep{Shih+2022}
    %         \item Kiyan and Adrian
    %         \item Matched Filter
    %         \item Gaia stuff: Adrian and Ana and Sara et al
    %         \item Uniform Modeling of 12 Stellar stream \citep{Patrick+2022}.
    %         Only on phi1, phi2, and CMD. Their CMD method is more sophisticated.
    %         Worth discussing.
    %     \end{enumerate}

    % Review of available data:
    % \begin{enumerate}
    %     \item Gaia. Streams in Gaia have on-sky positions and proper motions,
    %     and parallaxes, though for most streams the parallax errors are
    %     prohibitively large. Likewise, most streams in Gaia do not have good
    %     radial velocity measurements. The photometry is generally poor,
    %     generally necessitating cross-matching with other datasets.
    %     \item S5. This is a much smaller dataset. It has deeper photometry and
    %     radial velocity measurements.
    % \end{enumerate}

@@

        The paper is organized as follows.  In \autoref{sec:method} we introduce
        our method for characterizing stream tracks and computing stellar
        membership.  We discuss the data in
        \autoref{sub:method:framing_the_data} and build a probabilistic
        framework to model the stream and background in
        \autoref{sub:method:likelihood_setup}-\autoref{sub:methods:priors}.  In
        \autoref{sec:results_mock} we apply our method to simulated data to
        demonstrate the model with a known ground truth dataset, then to the
        stellar streams \stream{GD-1} (\autoref{sec:results_gd1}) and
        \stream{Pal\,5} (\autoref{sec:results_pal5}).  We compare our approach
        to characterizing stellar streams to other methods in
        \autoref{sec:comparison}.  We conclude in \autoref{sec:conclusions},
        discussing the results of this work and future directions.

@@

    We now introduce our method for characterizing stellar stream tracks and
    computing stream membership likelihoods from N-D phase-space measurements.
    % TODO: rephrase following sentence
    We develop a flexible Bayesian model framework, that allows arbitrary
    numbers of phase-space dimensions to be incorporated in the model and
    permits missing phase-space observations.  Below, we offer a concise
    overview of our approach, followed by a more in-depth technical explanation
    in subsequent sections.  We start, in \autoref{sub:method:framing_the_data},
    by transforming the astrometric data into a stream-oriented on-sky reference
    frame ($\phi_1, \phi_2, \parallax, \mu_{\phi_1}^*, \mu_{\phi_2}, v_r)$, for
    each stream. In this frame, the stream track lies along $\phi_1$, and the
    stream density may be described by distributions in the other parameters as
    injective functions of $\phi_1$.
    % TODO: ⬆ I think a few more sentences explaining why this is important,
    % feature variations, etc?  However, previously we discussed how stream will
    % vary along the track, with changes in the width and possible gaps.  Also
    % observational effects, e.g the distance gradient changes the on-sky width
    % (should this be mentioned earlier?)
    Likewise, the background of the Galactic field may be described with
    $\phi_1$-conditioned distributions.
    % TODO: ⬆ I think this should be expanded.
    Non-astrometric coordinates, e.g. photometry and metallicity, are also
    modelled along the stream and background as functions of $\phi_1$.

    The distributions for the stream and background are collected in a single
    mixture model that describes the entire field.  We take the parameters of
    the mixture model -- the mixing coefficients and component distribution
    parameters -- to be general (continuous) functions of the conditioning
    variable, $\phi_1$. This is done using feed-forward neural network that
    takes the conditioning variable as its input and output the parameter value.
    Mixture models of conditional probability distributions using neural
    networks are called Mixture Density Networks (MDNs). Since the model
    coefficients may evolve with $\phi_1$, MNDs are well suited to characterize
    streams (and the background) with all their variations in position, width,
    and linear density.

@@

        As a stream progenitor orbits the Galactic center of mass (GCOM), its
        orbital path describes an ellipse in the orbital plane. The associated
        arms of the stream generally align closely with its orbital plane.  From
        a Galactocentric reference frame, an orbital plane assumes the form of a
        great circle in the observer's celestial ``sky," thereby allowing the
        potential realignment of longitudinal ($\ell$) and latitudinal ($b$)
        coordinates to a coordinate system aligned with the stream
        ($\cmp{GCOM}{\phi_1}, \phi_2^{(GCOM)}$). Here, the progenitor is
        established at the origin, with the stream progressing along
        $\cmp{GCOM}{\phi_1} (\phi_2^{(GCOM)} \approx 0)$. $\cmp{GCOM}{\phi_1}$
        is clearly a good affine parameter along which to construct models and
        analyze the stream's characteristics.

        However, data observations are not taken from the GCOM, but from  $\sim
        8$ kpc distant \citep{GRAVITY2018, Leung+2022} in a local frame,
        typically from vantage points such as Earth or its solar Lagrange
        points, and subsequently represented in the International Celestial
        Reference System (ICRS) \citep{ICRS1997}.  Restricted as we are to a
        very local volume, there is no means to observe systems from the more
        ``intrinsic" reference frames. In our observational frames there is no
        inherent assurance of effecting a rotation that would similarly align
        the stream within observer coordinates such that $\phi_2(\phi_1) \approx
        0$.  Thankfully, most streams are adequately distant from both the
        Galactic center and the observer, or exhibit orientations such that a
        rotation a) can be established or b) holds valid. For most of the
        best-studied globular cluster streams -- \stream{GD-1}, \stream{Palomar
        5}, \stream{Jhelum}, etc. -- this holds true. However for some streams
        of dwarf galaxies, like the Magellanic stream which wraps multiple times
        around the Milky Way \citep{WannierWrixon1972}, no such transformation
        is possible. More generally, a stream's ``sky"-projected path may be
        kinked, e.g. by subhalo interactions, such that it is a many-to-one
        function in $\phi_1$.

        In this work we treat with the vast majority of streams, which have a
        ``sky" transformation such that the stream is an injective function in
        $\phi_2(\phi_1)$.  Since we are examining known streams this
        transformation is known \textit{a priori}, or may be constructed during
        pre-processing \citep[see][\S2.1]{Starkman+2023}.  We note that, with
        sufficient prior knowledge, many-to-one functions may be broken into
        injective segments and their connection constrained by a prior. Thus the
        methods developed subsequently may extented and applied even to wrapped
        and kinked streams. However, we leave extending the model framework to a
        future work.

@@

        \begin{figure}
            \centering
            \includegraphics[scale=0.5]{static/model/DAG_StreamDensity.pdf}
            \caption{Probabilistic Graphical Model of the Mixture Density Network (MDN)}
            \todo[inline]{TODO:}
            \begin{enumerate}
                \item remake this example this with
                \url{https://github.com/daft-dev/daft}. See
                \url{https://en.wikipedia.org/wiki/Mixture_model} for examples.
            \end{enumerate}
            \label{fig:DAG}
        \end{figure}

@@

    \subsection{Missing Phase-Space Observations}
    \label{sub:method:missing_data}

>>>>

\documentclass[twocolumn]{aastex631}
\usepackage{showyourwork}
\usepackage{amsmath, amssymb}
\usepackage{gensymb}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{dsfont}
\usepackage[inline]{enumitem}
\usepackage{cleveref}
\usepackage{transparent}
\usepackage{todonotes}
\usepackage{array}
\usepackage{float}

% Change the section names
\renewcommand{\sectionautorefname}{\S\!}%
\renewcommand{\subsectionautorefname}{\S\!\!}%
\renewcommand{\subsubsectionautorefname}{\S\!}%
\renewcommand{\equationautorefname}{Eq.}
\renewcommand{\tableautorefname}{Table}
\renewcommand{\figureautorefname}{Fig.}

\newcommand{\code}[1]{\textsc{#1}}
\newcommand{\package}[1]{\code{#1}}
\newcommand{\stream}[1]{#1}
\newcommand{\dataarchive}[1]{\textit{#1}}
\newcommand{\Gaia}{\dataarchive{Gaia}}
\newcommand{\PanStarrs}{\dataarchive{Pan-STARRS1}}

@@

% Statistics
\newcommand{\pdf}{p}
\newcommand{\cdf}{P}
\newcommand{\likelihood}{\mcal{L}}
\newcommand{\prior}{\mcal{\pi}}
\newcommand{\posterior}{p}

@@

\newcommand{\galaxy}[1]{
    \ifthenelse{\equal{#1}{}}{
        \includegraphics[width=9pt]{static/galaxy.png}
    }{
        \includegraphics[width={#1}pt]{static/galaxy.png}
    }%
}

\defcitealias{Ibata+2021}{I21}

@@

% Title
\title{Does this Star Belong? Data-Driven Characterization of Stellar Streams \\ with Mixture Density Networks}

% Author list
\author[0000-0003-3954-3291]{Nathaniel Starkman}
\affiliation{David A Dunlap Department of Astronomy and Astrophysics, University of Toronto}

\author[0000-0001-8042-5794]{Jacob Nibauer}
\altaffiliation{Co-first author.}
\affiliation{Department of Astrophysical Sciences, Princeton University}

\author[0000-0001-6855-442X]{Jo Bovy}
\affiliation{David A Dunlap Department of Astronomy and Astrophysics, University of Toronto}

\author[0000-0003-3613-0854]{Jeremy J. Webb}
\affiliation{Division of Natural Science, Department of Science, Technology and Society, York University}

% \author[0000-0002-7846-9787]{Ana Bonaca}
% \affiliation{Carnegie Observatories: Pasadena, California, US}

\correspondingauthor{Nathaniel Starkman}
\email{n.starkman@mail.utoronto.ca}

\shortauthors{Starkman and Nibauer et al.}

% Abstract with filler text
\begin{abstract}
    Stellar streams are sensitive probes of the Milky Way's gravitational
    potential. The mean track of a stream constrains global properties of the
    potential, while its fine-grained surface density contains information about
    galactic substructure. A precise characterization of streams from potentially noisy data marks a crucial step in constraining galactic structure across orders of magnitude in mass scales. Here we present a new method for constructing a smooth probability
    density model of stellar streams using all of the available astrometric and
    photometric data. To characterize the stream's morphology and kinematics, we
    utilize mixture density networks to construct a flexible representation of
    the stream and its on-sky track, width, stellar number density, and
    kinematic distribution. In photometric coordinates we model each stream as a
    single-stellar population, with a distance track that is simultaneously
    estimated from the stream's inferred distance modulus (using photometry) and
    parallax distribution (using astrometry). Normalizing flows are used to characterize the distribution of background stars. We apply the method to the stream
    \stream{GD-1}, and the tidal tails of \stream{Palomar 5}. For both streams we obtain a catalog
    of stellar membership probabilities that are made publicly available.
    Importantly, our model is capable of handling data with incomplete
    phase-space observations, making our method applicable to the growing census
    of Milky Way stellar streams. When applied to a population of streams, the
    resulting membership probabilities from our model form the required input to
    infer the Milky Way's dark matter distribution from the scale of the stellar
    halo down to subhalos. 
\end{abstract}

@@

    Stellar streams are the disrupted remnants of globular clusters and
    satellite galaxies, and their census has grown extensively in the Milky Way
    thanks to all-sky astrometric missions like \Gaia{}
    \citep{GaiaCollaboration+2016, GaiaCollaboration+2023} and low-surface
    brightness photometric surveys (e.g., \citealt{Ibata+1998,
    GrillmairDionatos2006, Shipp+2018, Ibata+2021}).  The formation of streams
    can be traced back to a progenitor (a globular cluster or satellite galaxy)
    whose stars are gradually lost to the tidal field of the host galaxy. The
    stars extend along a series of similar orbits, called tidal tails, which can
    sometimes span several tens of degrees across the sky. The orbital proximity
    of stars belonging to a stream makes them sensitive tracers of the mass
    distribution of galaxies, sourced by both the baryonic and dark matter
    components.

    The average properties of a stream (e.g., the mean position and velocity
    track of the tidal tails) can be used to reconstruct the global dark matter
    distribution of the stellar halo (e.g., \citealt{Johnston+1999, Bonaca+2014,
    Bovy2014, Bovy+2016, Nibauer+2022, Koposov+2023}), while local variations in
    the stream track and surface density encode information about the stream's
    encounters with baryonic structures such as the stellar bar (e.g.,
    \citealt{Erkal+2017, Pearson+2017}), and non-luminous structures like dark
    matter subhalos (e.g., \citealt{Ibata+2002, Johnston+2002,
    Carlberg+2012-74820C, Bovy+2017, Bonaca+2019, Hermans+2021}).

    While several methods have been developed to model stream properties as a
    function of global halo characteristics (e.g., the mass distribution,
    flattening; \citealt{Johnston+1999,2008MNRAS.386L..47B,Koposov+2010,2013MNRAS.433.1826S,Bovy2014,Nibauer+2022}) and local substructures (e.g., the bar's pattern
    speed and the mass of subhalo perturbers; \citealt{2014A&A...563A..60A,Pearson+2017,2016MNRAS.460..497H,Bovy+2017,Bonaca+2019}), these methods often
    rely on the existence of a homogenized stream-tracer population that
    characterizes a given stream across each of the observable phase-space
    dimensions. Additionally, because even the fine-grained properties of
    streams are influenced by the detailed mass distribution of its host galaxy,
    it is important to propagate errors both in measurement uncertainty and
    membership uncertainty (i.e., which stars belong to the stream versus the
    background) when attempting to model a stream. Otherwise, a poorly
    characterized stream could bias constraints on, e.g., the total mass of the
    galaxy, the dark matter halo shape, and the mass of subhalo perturbers. In
    order to generate more precise constraints on the structure of the galaxy,
    it is therefore imperative that a statistically sound and homogenized
    catalog of stellar streams is produced.

    While streams are well described in action-angle coordinates
    \citep[e.g.][]{Bovy2014} or (restricted) $N$-body simulations
    \citep[e.g.][]{Dehnen+2004}, these methods rely on prior assumptions about
    the galaxy and its underlying gravitational potential. Therefore, it is
    necessary to devise data-driven methods to characterize stellar streams
    without appealing to strong prior assumptions about the galaxy's structure.

    There are a few previous works which have modeled stellar streams in this
    context. \citet{Patrick+2022} developed the method introduced by
    \citet{Erkal+2017}, which is a data-driven, spline-based method for modeling
    the photometric properties and on-sky positions of previously identified
    streams. Their method works by fitting the stream's stellar surface density
    with a series of splines, capturing both surface brightness variations along
    the stream and changes in the position of the stream track and width. Their
    method also utilizes an isochrone fit to a given stream, from which a
    distance modulus can be derived. While this method is very successful at
    characterizing the photometric properties of a stream, it does not consider
    the kinematic dimensions which have become increasingly well-measured with
    successive data releases from \Gaia{}, as well as targeted observations.
    From a population of RR Lyrae stars, \citet{Price-Whelan+2019} utilized a
    mixture model to fit the globular cluster stream Pal-5 in the space of
    proper motions and heliocentric distances. The mixture modeling approach
    enables a simultaneous fit to both the stream and background density,
    allowing for better separation of the stream from the background
    distribution. However, the track of the stream in on-sky coordinates is not
    fit. The method identified 27 RR Lyrae stars consistent with being members
    of the stream. While useful for determining the average properties of a
    stream (e.g., its mean distance track), a limited population of RR Lyrae
    stars are not sufficient to capture the small-scale properties of streams
    expected to inform constraints on, e.g., a population of perturbers in the
    galaxy.

    Other methods for extracting stellar streams from the data include
    matched-filter algorithms, where over-densities in color-magnitude space are
    inspected as possible clusters undergoing tidal stripping
    \citep{Grillmair+1995, GrillmairJohnson2006, Shipp+2018}. While the
    matched-filter technique has been extremely successful in stream discovery,
    it is not equipped to characterize streams at the level of detail necessary
    to inform precise constraints on the potential of the galaxy and its
    substructure.

    In order to address the need for a flexible model of stellar streams both in
    photometric and kinematic spaces, in this work we develop a new method for
    characterizing streams and quantifying the membership probability of
    possible stream stars. Our approach provides several key advantage over
    previous stream modeling efforts, specifically in its ability to jointly
    model the kinematic and photometric properties of streams simultaneously
    using a flexible model.  Furthermore, our method is well-suited to fitting
    streams with only partial or incomplete phase-space or photometric
    observations, enabling a full characterization of streams insofar as the
    available data allows.

@@

        The paper is organized as follows.  In \autoref{sec:method} we introduce
        our method for characterizing stream tracks and computing stellar
        membership.  We discuss initial data processing in
        \autoref{sub:method:framing_the_data} and build a probabilistic
        framework to model the stream and background in
        \autoref{sub:method:likelihood_setup}-\autoref{sub:methods:priors}.  In
        \autoref{sec:results_mock} we apply our method to simulated data to
        demonstrate the model with a known ground truth dataset, then to the
        stellar streams \stream{GD-1} (\autoref{sec:results_gd1}) and
        \stream{Pal\,5} (\autoref{sec:results_pal5}).  We compare our approach
        to characterizing stellar streams to other methods in
        \autoref{sec:discussion}.  We conclude in \autoref{sec:conclusions},
        discussing the results of this work and future directions.

@@

    To characterize stellar stream tracks and for computing membership
    probabilities from kinematic and photometric measurements of stars, we
    develop a flexible Bayesian modeling framework that incorporates all of the
    available kinematic information for a field of stars, including those with
    only partial phase-space measurements. Below, we offer a concise overview of
    our approach, followed by a more in-depth technical explanation in
    subsequent sections.  We start, in \autoref{sub:method:framing_the_data}, by
    transforming the astrometric data into a stream-oriented on-sky reference
    frame ($\phi_1, \phi_2, \parallax, \mu_{\phi_1}^*, \mu_{\phi_2}, v_r)$, for
    each stream. In this frame, the stream track lies along $\phi_1$, and the
    stream density may be described by distributions in the other parameters as
    injective functions of $\phi_1$.  Likewise, the background of the Galactic
    field may be described with $\phi_1$-conditioned distributions.  We will
    also model the stream in photometric coordinates. Additional features, e.g.
    metallicity, can also be modeled along the stream and background as
    functions of $\phi_1$, though we leave this to a future work.

    The distributions for the stream and background are collected in a single
    mixture model that describes the entire field.  We take the parameters of
    the mixture model -- the mixture weights and component distribution
    parameters -- to be general (continuous) functions of the conditioning
    variable, $\phi_1$. This is done using a feed-forward neural network that
    takes the conditioning variable as its input and outputs the parameter
    value(s) that characterize the distribution of stars at a $\phi_1$ ``slice".
    Mixture models of conditional probability distributions using neural
    networks are called Mixture Density Networks (MDNs) \citep{Bishop+1994}.
    Since the model coefficients may evolve with $\phi_1$, MNDs are well suited
    to characterize streams (and the background) with all their variations in
    position, width, and linear density.

@@

        As a stream progenitor orbits the Galactic center of mass, its path can
        be approximated locally as an elliptical segment. The associated arms of
        the stream generally align closely with its orbital plane.  As viewed
        from a Galactocentric reference frame, the stream approximately assumes
        the form of a great circle enabling a 1:1 mapping from some phase
        parameter to a unique position along the stream. Thus in a `stream'
        oriented frame the rotated longitudinal and latitudinal coordinates
        ($\phi_1, \phi_2$) align with the stream such that the stream's phase is
        $\phi_1$ and $\phi_2 \simeq 0$.

        While it would be ideal to characterize streams in this Galactocentirc
        frame, transformation to the frame requires knowledge of the distance to
        the Galactic center  \citep{GRAVITY2018, BennettBovy2019, Leung+2022},
        the solar motion around the Galactic center, and accurate distances to
        field stars (among other uncertain quantities). In a heliocentric frame,
        e.g. the standard ICRS\footnote{International Celestial Reference
        System} \citep{ICRS1997}, projection effects can obscure the shape of a
        stream leading to a non-trivial stream track. Therefore in a
        heliocentric frame, there is no assurance of effecting a rotation like
        in the Galactocentric frame that would similarly align the stream in an
        observer's coordinates such that $\phi_2(\phi_1) \approx 0$.
        Thankfully, most streams are adequately distant from both the Galactic
        center and the observer, or exhibit orientations such that a rotation a)
        can be established or b) holds valid. For most of the best-studied
        globular cluster streams -- \stream{GD-1}, \stream{Palomar 5},
        \stream{Jhelum}, etc. -- this holds true. However for some streams of
        dwarf galaxies, like the Magellanic stream which wraps multiple times
        around the Milky Way \citep{WannierWrixon1972}, no such transformation
        is possible. More generally, a stream's ``sky"-projected path may be
        kinked, e.g. by subhalo interactions, such that it is a many-to-one
        function in $\phi_1$.

        In this work we treat with the vast majority of streams, which have a
        ``sky" transformation such that the stream is an injective function in
        $\phi_2(\phi_1)$.  Since we are examining known streams this
        transformation is known \textit{a priori}, or may be constructed during
        pre-processing \citep[see][\S2.1]{Starkman+2023}.  We note that, with
        sufficient prior knowledge, many-to-one functions may be broken into
        injective segments and their connection constrained by a prior. Thus the
        methods developed subsequently may be extended and applied even to
        wrapped and kinked streams. However, we leave extending the model
        framework to a future work.

@@

        \begin{figure*}
            \script{pgm.py}
            \centering
            \includegraphics[scale=0.6]{figures/pgm.pdf}
            \caption{%
                Probabilistic Graphical Models (PGMs) of the Mixture Density
                Networks used in this work.
                %
                \textbf{Left}: A PGM of the stream model. The full dataset
                $x_n^{obs}$ is made of the observed astrometry $w_n^{obs}$ and
                photometry $m_n^{obs}$, along with the corresponding
                observational errors. We model the data as a mixture of
                Gaussians (in astrometry + distance modulus), with the Gaussians
                indexed by $q$. The weight $\alpha$ determines the weighted
                contribution of each Gaussian to each datum. All models are
                conditioned on $\phi_1$.
                %
                \textbf{Right}: a PGM of the background model. The data is
                identical to that of the stream, however the model is not a
                mixture of Gaussians, but of a variety of distributions. We
                further distinguish between trainable distribution parameters
                (circles) and fixed distributions (points), like a pre-trained
                normalizing flow background model, explained in further detail
                in \autoref{sub:method:pre-training_distributions}. Alike to the
                stream model, all background models are conditioned on $\phi_1$.
                % 
            }
            \label{fig:PGM}
        \end{figure*}

@@

    %\newpage
    \subsection{Missing Phase-Space Observations}
    \label{sub:method:missing_data}

        Streams have a diversity of measurement coverage: several have full 6-D
        astrometrics as well as photometry and chemical abundances available
        \citep[e.g.,][]{Koposov+2019, Antoja+2020, Li+2022}, while most have
        partial coverage, lacking some feature dimensions.  Most streams in
        \Gaia \citep{Gaia2016, Gaia2023}, for instance, have proper motions but
        not radial velocities nor high signal-to-noise parallax measurements.
        Follow-up surveys, like \dataarchive{S5} \citep{Li+2019}, add many
        feature dimensions, however only for a select set of streams. Given the
        diversity of coverage it is important that the likelihood model --
        \autoref{eq:membership_prob} -- account for missing data.
